---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    theme: default
    keep_md: yes
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setoptions, echo=FALSE, results='hide', message=FALSE}
# Install required libraries and set global markdown options
library(tidyverse)
library(knitr)
opts_chunk$set(echo=FALSE, results="asis", 
               cache=TRUE, warning=TRUE, message=FALSE)
```


## Loading and preprocessing the data


This section describes the process to unzip dataset file and read it into a dataframe called "**activity**." The first two lines of code which have been commented out shows the steps to download the dataset if unavailable in the project folder. 


```{r echo=TRUE, results='hide'}
# file.URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
# download.file(file.URL, file.dir, method = "auto")


file.name <- "activity.csv"
file.dir <- "./activity.zip"
unzip(file.dir)
activity <- read.csv(file.name, na.strings = "NA")
file.remove(file.name)
```


Next, we will check the data and perform any pre-processing as required.
```{r}
# check dataframe
str(activity)


# check for missing values
colSums(is.na(activity))


# convert date from character to date format
activity$date <- as.Date(activity$date, format="%Y-%m-%d")
head(activity)
str(activity)
```





## What is mean total number of steps taken per day?





## What is the average daily activity pattern?





## Imputing missing values





## Are there differences in activity patterns between weekdays and weekends?





